Evidence accumulation modelling offers new insights into the cognitive mechanisms that underlie linguistic and action-based training

### 2024-11-26 manuscript/manuscript.tex
initial commit

---
title             : "Evidence accumulation modelling offers new insights into the cognitive mechanisms that underlie linguistic and action-based training"
shorttitle        : "Decision models and experience-dependent plasticity"
author: 
  - name          : "Samantha Parker"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "School of Psychological Sciences, 16 University Avenue, Macquarie University, NSW 2109"
    email         : "samantha.parker@students.mq.edu.au"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Data curation"
      - "Formal analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Emily S. Cross"
    affiliation   : "2"
    role:
      - "Conceptualization"
      - "Resources"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "2"
    role:
      - "Conceptualization"
      - "Visualisation"
      - "Supervision"
      - "Project Administration"
      - "Validation"
      - "Writing - Review & Editing"
affiliation:
  - id            : "1"
    institution   : "Macquarie University"
  - id            : "2"
    institution   : "ETH Zürich"
authornote: |
  Samantha Parker. School of Psychological Sciences, Macquarie University.
  Emily S. Cross. Social Brain Sciences Lab, Department of Humanities, Social and Political Sciences (D-GESS), ETH Zürich.
  Richard Ramsey. Neural Control of Movement Lab, Department of Health Sciences and Technology (D-HEST) and Social Brain Sciences Lab, Department of Humanities, Social and Political Sciences (D-GESS), ETH Zürich.
  <!-- Enter author note here. -->
abstract: |
  Evidence accumulation modelling has been shown to uncover new insights into the cognitive mechanisms that underlie decision making from behavioural data. By jointly modelling reaction time and accuracy data, such decision models estimate latent variables that represent distinct computaitonal processes, such as stimulus encoding, response caution and the quality of information processing. In this study we use an evidence accumulation model, the Linear Ballistic Accumulator (LBA), to shed new light into the mechanisms that underlie learning based on linguistic and action-based training. The LBA model is applied to behavioural data from a previously published training study where participants learn to name, tie or name and tie a set of knots. Our results show that training is multifaceted and associated with an increase in stimulus-encoding time, a reduction in response caution, as well as an increase in the speed of information accumulation. Furthermore, the results showed that there was an added benefit to the rate of evidence accumulation when naming and tying experience were combined. This latter finding suggests that performance benefits from multi-modal training may be instantiated in computational processes that are associated with the quantity and quality of information accumulation during decision making. Overall, in applying this computational approach to accuracy and reaction time data, we uncover new insights into the mechanisms that govern experience-dependent plasticity.

  *Public significance statement:* The results of this study show that multi-modal training, which spans linguistic and action-based content, increases the quality of information that is gained through perception to make a decision. These results provide a novel computational account of performance benefits that are often observed from multi-modal training, which will aid theory development on learning and plasticity within cognitive science.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "evidence accumulation modelling, training, learning, action, naming, cognitive mechanisms, plasticity"
wordcount         : "6703"
bibliography      : ["references.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
note              : "\\clearpage" ## this moves the author note section to a second page
editor_options: 
  chunk_output_type: console
header-includes:
  - \raggedbottom
---

```{r, include=FALSE}
## just include when de-bugging tinytex
# options(tinytex.verbose = TRUE)
```

```{r setup, include = FALSE}
pkg <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

lapply(pkg, library, character.only = TRUE)

r_refs(file = "r-references.bib")
my_r_citation <- cite_r(file = "r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

```{r plot-settings}
## theme settings for ggplot
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)

## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

```{r global-chunk-settings}
## set the figure options
knitr::opts_chunk$set(fig.pos='H', out.extra = '', out.height = "67%",
                      fig.align = "center") # initial version
```

# Introduction

Evidence accumulation models are a class of computational models used to understand the latent cognitive processes that underlie human decision making. Typically, these models use accuracy and reaction times collected in speeded choice tasks to draw insights about the psychological mechanisms that underlie those decisions. The application of this modelling approach within the domain of cognitive psychology has led to a number of novel insights about the mechanisms that underlie processes such as lexical decision making [@wagenmakers2008], ageing [@ratcliff2001; @ratcliff2010] and perceptual discrimination [@ratcliff2008]. Importantly, these insights are often not evident from more conventional analyses of accuracy and reaction time data. Given these characteristics, we applied an evidence accumulation model, the Linear Ballistic Accumulator (LBA; @brown2008), to behavioural data from a prior published study that used neuroimaging to examine the effects of different types of training on object knowledge and perception [@cross2012d]. In contrast to the original study’s analysis, which analysed accuracy and reaction time separately, in this paper we apply an evidence accumulation model to uncover novel insights into the latent computational processes that underlie linguistic and action-based learning.

Whilst it is common for researchers to draw conclusions from a separate analysis of accuracy and reaction time, it is unclear how to combine these measures into a single measure by which to quantify task difficulty or subject ability [@wagenmakers2007]. Consider, for example, a perceptual discrimination task that requires participants to indicate whether two images are of the same or different objects [@cross2012d; @weisberg2007]. Objects could belong to two categories of training, one where the participants learn to name the object (name-based) and the other where the participant interacts physically with the object (action-based). Objects that are subject to name-based training are discriminated faster but with more errors, than objects that are the subject of action-based training. It is unclear from a separate analysis of accuracy and reaction time data in which condition performance is superior. This is because without a principled way to combine accuracy and RT, these measures are incommensurable. While this example demonstrates the well-documented speed-accuracy trade-off, where less cautious responding is associated with faster, but more error prone responding, and more cautious responding is associated with slower but more accurate responses, this is just one of the many ways in which accuracy and reaction times can interact [@ratcliff1998].

Evidence accumulation models provide a principled way to combine accuracy and reaction time, enabling direct insight into the processes that underlie speeded decisions [@brown2008; @ratcliff2002]. They do so by combining accuracy and the distribution of RT for correct and error responses in order to estimate parameters of a model that can separate the effects of response caution from difficulty. While there are many different varieties of evidence accumulation model, they all share a similar basic framework [@donkin2009]. Namely, these models all assume that when making a decision, evidence is sampled from the environment and that information is used as evidence for one of the potential responses. As soon as evidence in favour of a particular response reaches a threshold, the decision process is terminated and the corresponding response is made.

All evidence accumulation models aim to provide estimates of four common aspects of the decision making process: the rate at which evidence accumulates in favour of a decision (drift rate), how much evidence is required before a response is reached (threshold), the amount of evidence in favour of a response that exists at the outset of a decision (start point noise) and the amount of time it takes to complete all processes that are thought to fall outside the decision, including stimulus encoding and motor responding (non-decision time). These parameters quantify the latent variables of the decision-making process. Threshold, for example, measures the amount of evidence necessary to make a response. As threshold increases, responses take longer but are more likely to be correct. In this way, threshold can account for differences in response caution. Drift rate, on the other hand, is thought to reflect the signal-to-noise ratio of the stimulus. In this way, drift rate quantifies the deterministic component of the accumulation process. Drift rate, therefore, measures both the quantity and the quality of information in the system and can provide a direct measure of task difficult or subject ability [@lewandowsky2018].

Assessing how parameters vary as a function of experimental conditions has the potential to reveal more about the mechanisms that underlie the decision-making process than is apparent from an analysis of accuracy and RT alone [@dutilh2019; @donkin2009; @evans2019; @evans2020]. Perhaps one of the most persuasive examples of this approach comes from the ageing literature. One typical finding within the ageing research is that response times increase with increasing age, a finding that led to the conclusion that there is a cognitive decline associated with ageing [@salthouse1996]. Application of an evidence accumulation model to data from such tasks, however, reveals that rather than older adults performing more poorly on speeded response time tasks than younger adults, differences in responses are actually due to differences in response caution [@ratcliff2004]. That is, older adults are more cautious responders than younger adults, requiring a higher level of evidence to trigger a response. Importantly, this inference was not available from a traditional analysis of accuracy and RT. This is just one example in which evidence accumulation modelling has allowed researchers to draw inferences that were unavailable in a traditional analysis of behavioural data [@white2022]. Evidence accumulation modelling thus provides us with an opportunity to interrogate behavioural data to uncover novel insights into the latent variables that underlie decision making in a way that a traditional analysis cannot.

In the present paper, we use the evidence accumulation approach to shed new light on the mechanisms that underlie two kinds of learning based on different types of training experience. Specifically, we apply the Linear Ballistic Accumulator (LBA) (Brown & Heathcote, 2008) to behavioural data from an existing multi-session training dataset that taught participants how to tie and name novel knots (Cross et al., 2012). In making use of existing data not only do we provide novel insights into the mechanisms that underlie different types of training, but we also demonstrate how computational modelling can lead to conclusions that are not readily apparent from a traditional analysis of accuracy and RT alone.

In the domain of training, the most typical behavioural finding is that with increased repetition of a task response times decrease and accuracy increases [@simons2016; @vonbastian2022]. While this finding is most commonly interpreted as evidence of improved task performance, it provides few insights about the latent processes that underlie training-related changes. Evidence accumulation modelling of training effects, on the other hand, has shown that task repetition can influence the rate at which information accumulates, the amount of evidence needed to trigger a response and non-decision time [@dutilh2009; @dutilh2011; @reinhartz2023]. Studies that have used these computational modelling techniques to investigate how task repetition improves response times report an increase in the rate of information accumulation associated with repeated performance of the same speeded task [@dutilh2009; @dutilh2011; @liu2012; @ratcliff2006; @zhang2014].

These studies have also found that while the rate of information accumulation increases during practice, these effects are also associated with decreased boundary separation (Dutilh et al., 2011; Dutilh et al., 2009; Liu & Watanabe, 2012; Zhang & Rowe, 2014). That is, with increased repetition of a task, participants require less evidence to trigger a decision. This finding has led some authors to suggest that training effects may elicit a shift in response caution, leading people to prioritise speed over accuracy. The influence of training on non-decision time has less consistently been reported in the literature, with some studies suggesting that training decreases non-decision timing (Dutilh et al., 2011; Dutilh et al., 2009) and other studies reporting no influence on this parameter [@strobach2013]. Dutilh and colleagues (2011; 2009), for example, found that improved performance on a repeated lexical decision task was associated with faster drift rates, lower response caution and a decrease in non-decision time. This finding led the authors to conclude that training effects were multifaceted.

These conclusions, however, have been drawn from a limited number of perceptual discrimination tasks, with the vast majority of studies using either random dot motion [@cochrane2023] or lexical discrimination tasks (Dutilh et al., 2011). Similarly, these studies typically investigate training related changes that are induced by task repetition (Dutilh et al., 2011; Dutilh et al., 2009). In contrast, we applied evidence accumulation modelling to understand and compare the mechanisms that underlie different types of training during a complex real-word perceptual discrimination task.

@@

The aim of the present study was to investigate to what extent parameters in the LBA model vary by each type of training. Given previous studies have suggested that increased drift rate, decreased response caution and decreased non-decision time (Dutilh et al., 2011; Dutilh et al., 2009; Zhang & Rowe, 2014) are all associated with training-related changes, we hypothesised that improvements in perceptual discrimination due to training would be associated with (1) an increase in drift rate; (2) a decrease in response caution; (3) a decrease in non-decision time or (4) some combination of these factors. Furthermore, it was also possible that the extent to which these parameters varied by training type (naming, tying, naming & tying, untrained) may differ, although we had no specific predictions about the direction of this hypothesis.

# Methods

## Transparency and Openness Statement

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study [@simmons2012]. In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: [https://aspredicted.org/6ztj-45s5.pdf](https://aspredicted.org/6ztj-45s5.pdf)). We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

Importantly, given that we were interested in the mechanisms that underlie training vs. no-training, as well as comparisons between different types of training (linguistic vs. action vs. both), only the post-training session was necessary to capture and quantify these effects. Indeed, Cross and colleagues (2012), in their original study, included only results from the post-training session in their final analysis, as this session is thought to be more consistent and reflect the participants' learned state [@weisberg2007]. All other aspects of the pre-registration were unchanged. In addition, all of the analysis code used in this paper have been made available on the open science framework ([https://osf.io/ea6gk/](https://osf.io/ea6gk/). Given that the data were collected several years before 2012, it was not yet routine to request explicit consent to share data publicly. Therefore, the raw data are only available from the authors upon request. Since the current manuscript was written in R using the papaja() package [@R-papaja], it is also computationally reproducible and available via GitHub ([https://github.com/rich-ramsey/knot_dmc](https://github.com/rich-ramsey/knot_dmc)).

## Dataset

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{evidence accumulation modelling, training, learning, action, naming, cognitive mechanisms, plasticity\newline\indent Word count: 6679}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evidence accumulation modelling offers new insights into the cognitive mechanisms that underlie linguistic and action-based training},
  pdfauthor={Samantha Parker1, Emily S. Cross2, \& Richard Ramsey2},
  pdflang={en-EN},
  pdfkeywords={evidence accumulation modelling, training, learning, action, naming, cognitive mechanisms, plasticity},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Evidence accumulation modelling offers new insights into the cognitive mechanisms that underlie linguistic and action-based training}
\author{Samantha Parker\textsuperscript{1}, Emily S. Cross\textsuperscript{2}, \& Richard Ramsey\textsuperscript{2}}
\date{}

\shorttitle{Decision models and experience-dependent plasticity}

\authornote{

Samantha Parker. School of Psychological Sciences, Macquarie University.
Emily S. Cross. Social Brain Sciences Lab, Department of Humanities, Social and Political Sciences (D-GESS), ETH Zürich.
Richard Ramsey. Neural Control of Movement Lab, Department of Health Sciences and Technology (D-HEST) and Social Brain Sciences Lab, Department of Humanities, Social and Political Sciences (D-GESS), ETH Zürich.

The authors made the following contributions. Samantha Parker: Conceptualization, Data curation, Formal analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily S. Cross: Conceptualization, Resources, Writing - Review \& Editing; Richard Ramsey: Conceptualization, Visualisation, Supervision, Project Administration, Validation, Writing - Review \& Editing.

Correspondence concerning this article should be addressed to Samantha Parker, School of Psychological Sciences, 16 University Avenue, Macquarie University, NSW 2109. E-mail: \href{mailto:samantha.parker@students.mq.edu.au}{\nolinkurl{samantha.parker@students.mq.edu.au}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Macquarie University\\\textsuperscript{2} ETH Zürich}

\note{\clearpage}

\abstract{%
Evidence accumulation modelling has been shown to uncover new insights into the cognitive mechanisms that underlie decision making from behavioural data. By jointly modelling reaction time and accuracy data, such decision models estimate latent variables that represent distinct computaitonal processes, such as stimulus encoding, response caution and the quality of information processing. In this study we use an evidence accumulation model, the Linear Ballistic Accumulator (LBA), to shed new light into the mechanisms that underlie learning based on linguistic and action-based training. The LBA model is applied to behavioural data from a previously published training study where participants learn to name, tie or name and tie a set of knots. Our results show that training is multifaceted and associated with an increase in stimulus-encoding time, a reduction in response caution, as well as an increase in the speed of information accumulation. Furthermore, the results showed that there was an added benefit to the rate of evidence accumulation when naming and tying experience were combined. This latter finding suggests that performance benefits from multi-modal training may be instantiated in computational processes that are associated with the quantity and quality of information accumulation during decision making. Overall, in applying this computational approach to accuracy and reaction time data, we uncover new insights into the mechanisms that govern experience-dependent plasticity.

\emph{Public significance statement:} The results of this study show that multi-modal training, which spans linguistic and action-based content, increases the quality of information that is gained through perception to make a decision. These results provide a novel computational account of performance benefits that are often observed from multi-modal training, which will aid theory development on learning and plasticity within cognitive science.
}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Evidence accumulation models are a class of computational models used to understand the latent cognitive processes that underlie human decision making. Typically, these models use accuracy and reaction times collected in speeded choice tasks to draw insights about the psychological mechanisms that underlie those decisions. The application of this modelling approach within the domain of cognitive psychology has led to a number of novel insights about the mechanisms that underlie processes such as lexical decision making (Wagenmakers, Ratcliff, Gomez, \& McKoon, 2008), ageing (Ratcliff, Thapar, \& McKoon, 2001, 2010) and perceptual discrimination (Ratcliff \& McKoon, 2008). Importantly, these insights are often not evident from more conventional analyses of accuracy and reaction time data. Given these characteristics, we applied an evidence accumulation model, the Linear Ballistic Accumulator (LBA; Brown and Heathcote (2008)), to behavioural data from a prior published study that used neuroimaging to examine the effects of different types of training on object knowledge and perception (Cross et al., 2012). In contrast to the original study's analysis, which analysed accuracy and reaction time separately, in this paper we apply an evidence accumulation model to uncover novel insights into the latent computational processes that underlie linguistic and action-based learning.

Whilst it is common for researchers to draw conclusions from a separate analysis of accuracy and reaction time, it is unclear how to combine these measures into a single measure by which to quantify task difficulty or subject ability (Wagenmakers, Van Der Maas, \& Grasman, 2007). Consider, for example, a perceptual discrimination task that requires participants to indicate whether two images are of the same or different objects (Cross et al., 2012; Weisberg, van Turennout, \& Martin, 2007). Objects could belong to two categories of training, one where the participants learn to name the object (name-based) and the other where the participant interacts physically with the object (action-based). Objects that are subject to name-based training are discriminated faster but with more errors, than objects that are the subject of action-based training. It is unclear from a separate analysis of accuracy and reaction time data in which condition performance is superior. This is because without a principled way to combine accuracy and RT, these measures are incommensurable. While this example demonstrates the well-documented speed-accuracy trade-off, where less cautious responding is associated with faster, but more error prone responding, and more cautious responding is associated with slower but more accurate responses, this is just one of the many ways in which accuracy and reaction times can interact (Ratcliff \& Rouder, 1998).

Evidence accumulation models provide a principled way to combine accuracy and reaction time, enabling direct insight into the processes that underlie speeded decisions (Brown \& Heathcote, 2008; Ratcliff, 2002). They do so by combining accuracy and the distribution of RT for correct and error responses in order to estimate parameters of a model that can separate the effects of response caution from difficulty. While there are many different varieties of evidence accumulation model, they all share a similar basic framework (Donkin, Averell, Brown, \& Heathcote, 2009). Namely, these models all assume that when making a decision, evidence is sampled from the environment and that information is used as evidence for one of the potential responses. As soon as evidence in favour of a particular response reaches a threshold, the decision process is terminated and the corresponding response is made.

All evidence accumulation models aim to provide estimates of four common aspects of the decision making process: the rate at which evidence accumulates in favour of a decision (drift rate), how much evidence is required before a response is reached (threshold), the amount of evidence in favour of a response that exists at the outset of a decision (start point noise) and the amount of time it takes to complete all processes that are thought to fall outside the decision, including stimulus encoding and motor responding (non-decision time). These parameters quantify the latent variables of the decision-making process. Threshold, for example, measures the amount of evidence necessary to make a response. As threshold increases, responses take longer but are more likely to be correct. In this way, threshold can account for differences in response caution. Drift rate, on the other hand, is thought to reflect the signal-to-noise ratio of the stimulus. In this way, drift rate quantifies the deterministic component of the accumulation process. Drift rate, therefore, measures both the quantity and the quality of information in the system and can provide a direct measure of task difficult or subject ability (Lewandowsky \& Oberauer, 2018).

Assessing how parameters vary as a function of experimental conditions has the potential to reveal more about the mechanisms that underlie the decision-making process than is apparent from an analysis of accuracy and RT alone (Donkin et al., 2009; Dutilh et al., 2019; Evans, 2019; Evans \& Wagenmakers, 2020). Perhaps one of the most persuasive examples of this approach comes from the ageing literature. One typical finding within the ageing research is that response times increase with increasing age, a finding that led to the conclusion that there is a cognitive decline associated with ageing (Salthouse, 1996). Application of an evidence accumulation model to data from such tasks, however, reveals that rather than older adults performing more poorly on speeded response time tasks than younger adults, differences in responses are actually due to differences in response caution (Ratcliff, Thapar, Gomez, \& McKoon, 2004). That is, older adults are more cautious responders than younger adults, requiring a higher level of evidence to trigger a response. Importantly, this inference was not available from a traditional analysis of accuracy and RT. This is just one example in which evidence accumulation modelling has allowed researchers to draw inferences that were unavailable in a traditional analysis of behavioural data (White \& Kitchen, 2022). Evidence accumulation modelling thus provides us with an opportunity to interrogate behavioural data to uncover novel insights into the latent variables that underlie decision making in a way that a traditional analysis cannot.

In the present paper, we use the evidence accumulation approach to shed new light on the mechanisms that underlie two kinds of learning based on different types of training experience. Specifically, we apply the Linear Ballistic Accumulator (LBA) (Brown \& Heathcote, 2008) to behavioural data from an existing multi-session training dataset that taught participants how to tie and name novel knots (Cross et al., 2012). In making use of existing data not only do we provide novel insights into the mechanisms that underlie different types of training, but we also demonstrate how computational modelling can lead to conclusions that are not readily apparent from a traditional analysis of accuracy and RT alone.

In the domain of training, the most typical behavioural finding is that with increased repetition of a task response times decrease and accuracy increases (Simons et al., 2016; von Bastian et al., 2022). While this finding is most commonly interpreted as evidence of improved task performance, it provides few insights about the latent processes that underlie training-related changes. Evidence accumulation modelling of training effects, on the other hand, has shown that task repetition can influence the rate at which information accumulates, the amount of evidence needed to trigger a response and non-decision time (Dutilh, Krypotos, \& Wagenmakers, 2011; Dutilh, Vandekerckhove, Tuerlinckx, \& Wagenmakers, 2009; Reinhartz, Strobach, Jacobsen, \& von Bastian, 2023). Studies that have used these computational modelling techniques to investigate how task repetition improves response times report an increase in the rate of information accumulation associated with repeated performance of the same speeded task (Dutilh et al., 2011; Dutilh et al., 2009; Liu \& Watanabe, 2012; Ratcliff, Thapar, \& McKoon, 2006; Zhang \& Rowe, 2014).

These studies have also found that while the rate of information accumulation increases during practice, these effects are also associated with decreased boundary separation (Dutilh et al., 2011; Dutilh et al., 2009; Liu \& Watanabe, 2012; Zhang \& Rowe, 2014). That is, with increased repetition of a task, participants require less evidence to trigger a decision. This finding has led some authors to suggest that training effects may elicit a shift in response caution, leading people to prioritise speed over accuracy. The influence of training on non-decision time has less consistently been reported in the literature, with some studies suggesting that training decreases non-decision timing (Dutilh et al., 2011; Dutilh et al., 2009) and other studies reporting no influence on this parameter (Strobach, Liepelt, Pashler, Frensch, \& Schubert, 2013). Dutilh and colleagues (2011; 2009), for example, found that improved performance on a repeated lexical decision task was associated with faster drift rates, lower response caution and a decrease in non-decision time. This finding led the authors to conclude that training effects were multifaceted.

These conclusions, however, have been drawn from a limited number of perceptual discrimination tasks, with the vast majority of studies using either random dot motion (Cochrane, Sims, Bejjanki, Green, \& Bavelier, 2023) or lexical discrimination tasks (Dutilh et al., 2011). Similarly, these studies typically investigate training related changes that are induced by task repetition (Dutilh et al., 2011; Dutilh et al., 2009). In contrast, we applied evidence accumulation modelling to understand and compare the mechanisms that underlie different types of training during a complex real-word perceptual discrimination task.

@@

The aim of the present study was to investigate to what extent parameters in the LBA model vary by each type of training. Given previous studies have suggested that increased drift rate, decreased response caution and decreased non-decision time (Dutilh et al., 2011; Dutilh et al., 2009; Zhang \& Rowe, 2014) are all associated with training-related changes, we hypothesised that improvements in perceptual discrimination due to training would be associated with (1) an increase in drift rate; (2) a decrease in response caution; (3) a decrease in non-decision time or (4) some combination of these factors. Furthermore, it was also possible that the extent to which these parameters varied by training type (naming, tying, naming \& tying, untrained) may differ, although we had no specific predictions about the direction of this hypothesis.

\section{Methods}\label{methods}

\subsection{Transparency and Openness Statement}\label{transparency-and-openness-statement}

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study (Simmons, Nelson, \& Simonsohn, 2012). In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: \url{https://aspredicted.org/6ztj-45s5.pdf}). We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

Importantly, given that we were interested in the mechanisms that underlie training vs.~no-training, as well as comparisons between different types of training (linguistic vs.~action vs.~both), only the post-training session was necessary to capture and quantify these effects. Indeed, Cross and colleagues (2012), in their original study, included only results from the post-training session in their final analysis, as this session is thought to be more consistent and reflect the participants' learned state (Weisberg et al., 2007). All other aspects of the pre-registration were unchanged. In addition, all of the analysis code used in this paper have been made available on the open science framework (\url{https://osf.io/ea6gk/}. Given that the data were collected several years before 2012, it was not yet routine to request explicit consent to share data publicly. Therefore, the raw data are only available from the authors upon request. Since the current manuscript was written in R using the papaja() package (Aust \& Barth, 2023), it is also computationally reproducible and available via GitHub (\url{https://github.com/rich-ramsey/knot_dmc}).
